---
title: "BIOS 617 - Lecture 1"
author: "Walter Dempsey"
date: "1/8/2019"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## BIOS 617 slides in R Markdown

All course slides will be written as R Markdown presentations. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document. This lets me share slides and code all in one document making all lecture content reproducible!

PDF copies are available on **canvas** while the RMD files and data will be available at <https://github.com/wdempsey/BIOS617>.

## Probability sampling?

- Statisticians receive data to analyze
  - How is it collected?
  - Common assumptions
- Historical strategies involved non-probability sampling
  - Purposive samples:
    - Choose both average and extreme elements to cover range of distribution
    - Useful strategy for h
- No guarantee that any of these are *representative*

## Example 1: Brexit data
```{r brexit, echo = FALSE, include = FALSE}
# load libraries
library(dslabs)
library(tidyverse)
library(ggrepel)
library(matrixStats)

# set colorblind-friendly color palette
colorblind_palette <- c("black", "#E69F00", "#56B4E9", "#009E73",
                        "#CC79A7", "#F0E442", "#0072B2", "#D55E00")
data(brexit_polls)
```

```{r cars, echo = FALSE, fig.align='center'}
# plot of Brexit referendum polling spread between "Remain" and "Leave" over time
brexit_polls %>%
  ggplot(aes(enddate, spread, color = poll_type)) +
  geom_hline(aes(yintercept = -.038, color = "Actual spread")) +
  geom_smooth(method = "loess", span = 0.4) +
  geom_point() +
  scale_color_manual(values = colorblind_palette[1:3]) +
  xlab("Poll end date (2016)") +
  ylab("Spread (Proportion Remain - Proportion Leave)") +
  labs(color = "Poll type") +
  ggtitle("Spread of Brexit referendum online and telephone polls")
```

## Slide with Plot

```{r pressure}
plot(pressure)
```

