---
title: "BIOS 617 - Lecture 2"
author: "Walter Dempsey"
date: "1/13/2019"
output:
  ioslides_presentation:
    css: styles.css
  beamer_presentation: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## JITT Review/Discussion 

## Simple random sampling (SRS) without replacement

- Generate sample of size $n$ from $N$ by sampling without replacement
- Given $j$ *unique* selected units, the next unit is sampled uniformly at random from remaining $N-j$ units each with probability $1/(N-j)$.
- There are ${N \choose n} = \frac{N!}{n! (N-n)!}$ possible samples
- **Show**: sample mean under SRS is unbiased
- **Show**: sample variance is $(1-f) \frac{S^2}{n}$ 

## Unbiasedness of SRS sample mean

* What is the probability of sampling individual $j$?
$$\small
\begin{aligned}
P(I_j = 1) &= 1 - P(I_j \neq 1) \\
&= 1 - \frac{N-1}{N} \times \frac{N-2}{N-1} \times \cdots \times \frac{N-n}{N-(n-1)} \\
&= 1 - \frac{N-n}{N} = \frac{n}{N}
\end{aligned}
$$
* Then by linearity of expectations
$$\small
E [ \bar y ] = E \left[ \frac{1}{n} \sum_j I_j Y_j \right] = \frac{1}{n} \sum_j E [ I_j  ] Y_j = \bar Y
$$ 

## A slightly more cumbersome proof

We can use the the sampling design directly instead. The number of samples $S = {N \choose n}$.
Then

$$\small
\begin{aligned}
E[ \bar y ] &= \sum_{s=1}^S P_s \bar y_s = \sum_{s=1}^S {N \choose n}^{(-1)} \frac{1}{n} \sum_{j=1}^N I^{(s)}_j Y_j \\
 &= \frac{(N-n)!n!}{n \cdot N!} \sum_{j=1}^N Q_j Y_j = \frac{(N-n)!n!}{n \cdot N!} \sum_{j=1}^N {N-1 \choose n-1} Y_j \\
 &= \frac{(N-n)!n!}{n \cdot N!} \frac{(N-1)!}{(n-1)! (N-n)!} \sum_{j=1}^N Y_j \\
 &= \frac{1}{N} \sum_{j=1}^N Y_j \\
\end{aligned}
$$

## Variance of $\bar y$

$$ \small
\begin{aligned}
&E \left[ \left( \bar y - \bar Y \right)^2  \right] \\
= &E \left[ \left( n^{-1} \sum_{j=1}^N I_j^{(s)} (Y_i - \bar Y) \right)^2  \right] \\
= &\frac{1}{n^2} \left(  \sum_{i,j=1}^N E \left[I_i^{(s)} I_j^{(s)} \right] (Y_i - \bar Y) (Y_j - \bar Y)  \right) \\
= &\frac{1}{n^2}  \left( \sum_{i}^N \frac{n}{N} (Y_i - \bar Y)^2  + 
\sum_{i \neq j} \frac{n}{N} \cdot \frac{n-1}{N-1} (Y_i - \bar Y) (Y_j - \bar Y)  \right) \\
= &\frac{1}{n \cdot N}  \left( \left(1 - \frac{n-1}{N-1} \right) \sum_{i}^N (Y_i - \bar Y)^2  + 
\frac{n-1}{N-1}\left[ \sum_{i=1}^N  (Y_i - \bar Y) \right]^2  \right) \\
= &\frac{N-n}{N} \frac{S^2}{n} + 0 = (1-f)\frac{S^2}{n}
\end{aligned}
$$


## Method 2: Symmetry arguments

Let's use prior class results on variance directly 
$$ \small
\begin{aligned}
V(\bar y) = &V \left( n{-1} \sum_{i=1}^n y_i \right) = \frac{1}{n^2} V \left( \sum_{i=1}^n y_i \right) \\
&= \frac{1}{n^2} \left[ n \cdot V(y_i) + n \cdot (n-1) \cdot \text{Cov} (y_i, y_j) \right] \\ 
&= \frac{1}{n^2} \left[ n \cdot \frac{N-1}{N} S^2 + n \cdot (n-1) \left( E[y_i y_j] - \bar Y^2 \right) \right] 
\end{aligned}
$$

## Method 2: Computing moments
$$ \small
\begin{aligned}
E[y_i y_j ] &= E \left[ \frac{y_j}{N-1} \sum_{i=1, i \neq j}^N Y_i \right] = E \left[ \frac{y_j}{N-1} (N \bar Y - y_j) \right]\\
&= \frac{1}{N-1} \left( N \bar Y^2 - E [ y_j^2 ] \right) = \frac{1}{N-1} \left( N \bar Y^2 - \bar Y^2 - \frac{N-1}{N} S^2   \right) \\
&= \bar Y^2 - \frac{S^2}{N}
\end{aligned}
$$
So the $\text{Cov} (y_i, y_j) = - \frac{S^2}{N}$ and
$$
 V(\bar y) = \frac{1}{n} \left[ (N-1)\frac{S^2}{N} - (n-1) \frac{S^2}{N}  \right] = \frac{S^2}{n} (1-f).
$$

## Simulation study (ctd.)

```{r packages, echo = FALSE, include = FALSE}
# load libraries
library(dslabs)
library(tidyverse)
library(ggrepel)
library(matrixStats)
```


```{r heights, echo = FALSE, fig.align='center'}
set.seed(1)
N = 100
heights = rnorm(n = N, mean = 65, sd = 5)
heights = data.frame(heights)
heights %>%
  ggplot(aes(heights)) + geom_histogram(binwidth = 1) +  
  xlab("Student height (in inches)") +
  ylab("Count") +
  ggtitle("Simulated height of students in BIOS617")
```

## Design-based inference: Simple example
```{r sampled, echo = FALSE}
set.seed(179)
n = 10
swor = sample(1:35, size = n, replace = FALSE)
Y = heights$heights
y = heights$heights[swor]
swor_result = mean(y)
S = var(Y)
f = n/N
variance = (1-f)*S/n
est_variance = (1-f)*var(y)/n
```

* Population's average height is `r round(mean(heights$heights),4)`
* Population's $S^2$ is `r round(S,4)`
* Example sample w/o replacement: `r swor`
* Corresponding estimate: `r round(swor_result,4)`
* Variance is `r round(variance,4)`
* Estimated variance is `r round(est_variance,4)`

## Design-based inference: Simulation tests
```{r sampled2, echo = FALSE, include = FALSE}
set.seed(1651891)
n = 10; num.iters = 10000
swor_results = vector(length = num.iters)
var_swor_results = vector(length = num.iters)
for (i in 1:num.iters) {
  swor = sample(1:35, size = n, replace = FALSE)
  swor_results[i] = mean(heights$heights[swor])
  var_swor_results[i] = var(heights$heights[swor])
}
mean((swor_results - mean(swor_results))^2)
variance

```

* Population's average height is `r round(mean(heights$heights),4)`
* Population's $S$ is `r round(S,4)`
* Simulate 10,000 simple random samples
* Calculate $\bar y^{(k)}$ for the $k$th simulation ($k=1,\ldots, 10,000$)
* Compute empirical variance `r round(var(swor_results), 4)`
* Compare with $(1-f)*S^2/n$ is `r round(variance,4)`



## Sampling with replacement

- In practice, we almost never sample with replacement
- Often, variance estimators are easier to compute
- Usually only estimator with analytic solution 

## Unbiased estimation of $S^2$

- Variance $(1-f) \frac{S^2}{n}$ depends on unknown quantity $S^2$
- **Idea**: Use the sample variance as an estimator
- **Question**: Is this an unbiased estimator of $S^2$?
$$ \small
\begin{aligned}
E \left( s^2 \right) &= E \left( \frac{1}{n-1} \sum_{i=1}^n (y_i - \bar y)^2 \right) \\
&= E \left( \frac{1}{n-1} \sum_{i=1}^n ( (y_i - \bar Y) - (\bar y - \bar Y))^2 \right) \\
\end{aligned}
$$
- The cross-term is zero because
$$ \small
\sum_{i=1}^n (y_i - \bar y) (\bar y - \bar Y) =  (\bar y - \bar Y) \sum_{i=1}^n (y_i - \bar y) = 0
$$

## Unbiased estimation of $S^2$

$$ \small
\begin{aligned}
&E \left( \frac{1}{n-1} \sum_{i=1}^n ( (y_i - \bar Y)^2 - (\bar y - \bar Y)^2 \right) \\
=& \frac{n}{n-1} \times \frac{N-1}{N} S^2 - \frac{n}{n-1} \frac{1}{n} S^2 (1-f) \\
=& \frac{S^2}{n-1} \left( n \frac{N-1}{N} - \left(1-\frac{n}{N} \right) \right) \\
=& \frac{S^2}{n-1} \left( \frac{n \cdot(N-1)-(N-n)}{N} \right) \\
=& \frac{S^2}{n-1} \left( \frac{N(n-1)}{N} \right) = S^2.
\end{aligned}
$$

## Standard errors and confidence limits

Typically standard errors and confidence intervals are computed using a normal approximation:
$$ \small
\frac{\bar y_n - \bar Y_N}{\sqrt{\text{Var}(\bar y_n)}} \to N(0,1)
$$
in distribution as $n \to \infty$, $N \to \infty$, and $n/N \to 0$.

The proof for this is beyond the scope of this course; it is a more complex version of the standard central limit theorem (which assumes infinite populations). A key assumption is a variant on the existence of variance (Lindeberg condition [independence, not IID] extended to finite population sampling):

## Estimating proportions under SRS

* Let $Y_i = 1$ is element $i$ has a given attribute, and $0$ otherwise.
* $\bar Y = M/N=P$ where 
  + $M$ is the number of elements in the population with that attribute
  + $P$ is the proportion of the population with the attribute
- The variance is
$$ \small
\begin{aligned}
S^2 &= \frac{\sum_{i=1}^N (Y_i - \bar Y)^2}{N-1} = \frac{\sum_{i=1}^N Y_i^2 - N\bar Y^2)}{N-1} \\
&= \frac{NP - NP^2}{N-1} = \frac{N \cdot P \cdot (1-P)}{N-1}
\end{aligned}
$$

## Estimating proportions under SRS

* Sample proportion $p = n^{-1} \sum_{i=1}^n y_i$ unbiased for $P$
* Sample variance $\frac{n p (1-p)}{n-1}$ is unbiased for $S^2$
* $v(p) = (1-f) \frac{p (1-p)}{n-1}$ 
* As $N$ gets large $f \to 0$ and so, for large~$n$, 
  + $v(p) \to \frac{p(1-p)}{n}$ 
  + "Standard" variance for binomial variable under normal approximation

## Design-based inference: Simulation tests
* Suppose we want to know number of students taller than 62 inches
* The sample has `r sum(y > 62)` out of the `r n` individuals
* Population has `r mean(Y > 62)` fraction of individuals
* Sample variance is `r round(n * mean(y > 62) *(1-mean(y > 62))/(n-1),3)` 
* True variance is `r round(N * mean(Y > 62) *(1-mean(Y > 62))/(N-1),3)` 

## Subdomain estimation

- In some cases, we are interested in inference for a particular subdomain (say females, households with incomes below $30K/year, businesses of a certain size, etc.)

- Here the distinction is that we often cannot ascertain whether an
element in the population is a member of the subdomain until we
sample it.

- Let $j$ index the subdomain of interest, and $D_{ij}$ be an indicator that the $i$th element belongs to subdomain $j$. 

- Then sample estimator $\bar y_j = n_j^{-1} \sum_{i=1}^n D_{ij} y_i$ with $n_{j} = \sum_{i=1}^n D_{ij}$
is unbiased for the population subdomain $\bar Y_j = N_j^{-1} \sum_{i=1}^N D_{ij} Y_i$ for $N_j = \sum_{i=1}^N D_{ij}$

## Subdomain estimation

* The term $n_j$ is a random variable
* Variance of $\bar y_j$ dpeends on $n_j$, which is a random variable
* If we condition on $n_j$, then we have the obvious extensions:
$$ \small
V(\bar y_j) = (1-f) \frac{S_j^2}{n_j}, \text{ where } S_j^2 = (N_j - 1)^{-1} \sum_{i=1}^N D_{ij} (Y_i - \bar Y_j)^2
$$
* And
$$ \small
v(\bar y_j) = (1-f) \frac{s_j^2}{n_j}, \text{ where } s_j^2 = (n_j - 1)^{-1} \sum_{i=1}^n D_{ij} (y_i - \bar y_j)^2
$$
* If $N_j$ is known, then $f = n/N$ can be replaced by $f_j = n_j / N_j$

## Subdomain totals

* Unbiased estimator of the subdomain total $Y_j = \sum_{i=1}^N D_{ij} Y_i$ is 
$$\small 
y_j = \frac{N}{n} \sum_{i=1}^n D_{ij} y_i
$$
* Also, $v(y_j) = \frac{N^2}{n} \tilde s^2 (1-f)$ for 
$$\small
\tilde s^2 = (n-1)^{-1} \sum_{i=1}^n D_{ij} (y_i - \bar{\tilde y}_j)^2
$$
where $\bar{\tilde y}_j = n^{-1} \sum_{i=1}^n D_{ij} y_i$

## Your second JITT:  3rd proof of variance
- We know $V (\bar y) = V\left( \frac{1}{n} \sum_{i=1}^n I_i Y_i \right)$ 
- Hint 1: $V(I_i) = f (1-f)$ and $\text{Cov}(I_i, I_j) = f \left( \frac{n-1}{N-1} - f \right)$
- Step 1: Use hint 1 to show the variance is equivalent to
$$ \small
\frac{f (1-f)}{n^2} \left[ \sum_{i=1}^N Y_i^2 - \frac{1}{N-1} \sum_{j=1}^N \sum_{i\neq j} Y_i Y_j\right]
$$
- Hint 2:
$$ \small
\sum_{j=1}^N \sum_{i=1}^N Y_i Y_j = \left( \sum_{i=1}^N Y_i \right)^2 - \sum_{i=1}^N Y_i^2
$$

- Step 2: Combine hint and variance formula to show result
